---
title: "Data Analysis (refactored)"
output: html_document
date: "2025-03-17"
---

```{r}
db <- read.csv("data/wb-covid-data.csv")
db <- db[, -1]
```

``` {r, include = FALSE}
db
```


### Descriptive Plots

#### Focus: `CovidCases`

```{r}
par(mar = c(4, 5, 3, 1), cex.axis = 0.50)
boxplot(db$CovidCases ~ db$Govt, 
        main = "Boxplot of COVID-19 Cases by Government Type",
        xlab = "Government Type",
        ylab = "Number of Cases",
        col = "greenyellow")
par(mar = c(4, 5, 3, 1), cex.axis = 0.75)
hist(db$CovidCases, 
     main = "Histogram of COVID-19 Cases as of April 22, 2020", 
     xlab = "Total Cases", 
     ylab = "Number of Countries", 
     col = "coral1", 
     breaks = 20)
```

The boxplots indicate that the distribution of `CovidCases` for each factor of `Govt` seem to be heavily right-skewed, and the histogram of `CovidCases` is indicative of an approximately exponential distribution. This suggests a natural log transformation of `CovidCases`. A new feature `logCases` is created.

```{r}
par(mar = c(4, 5, 3, 1), cex.axis = 0.50)
boxplot(log(db$CovidCases) ~ db$Govt, 
        main = "Boxplot of (ln) COVID-19 Cases By Government Type", 
        xlab = "Government Type", 
        ylab = "(ln) Number of Cases", 
        col = "greenyellow")
db$logCases <- log(db$CovidCases)
```

---

#### Focus: `CovidDeaths`

```{r}
par(mar = c(4, 5, 3, 1), cex.axis = 0.50)
boxplot(db$CovidDeaths ~ db$Govt, 
        main = "Boxplot of COVID-19 Deaths By Government Type", 
        xlab = "Government Type", 
        ylab = "Number of Deaths", 
        col = "royalblue1")
par(mar = c(4, 5, 3, 1), cex.axis = 0.75)
hist(db$CovidDeaths, 
     main = "Histogram of COVID-19 Deaths as of April 22, 2020", 
     xlab = "Total Deaths", 
     ylab = "Number of Countries", 
     col = "darkslategray1", 
     breaks = 20)
```

```{r}
db$logDeaths <- log(db$CovidDeaths + 1)
par(mar = c(4, 5, 3, 1), cex.axis = 0.50)
boxplot(db$logDeaths ~ db$Govt, 
        main = "Boxplot of (ln) COVID-19 Deaths By Government Type",
        xlab = "Government Type", 
        ylab = "(ln) Number of Deaths", 
        col = "royalblue1")
```

A similar distribution for `CovidDeaths` is evident. Note: applying a natural log transformation to `CovidDeaths` results in a handful of "-infinity" values, due to the fact that some nations have a death count of 0. To remedy this, `CovidDeaths` is incremented by 1 before creating `logDeaths`.

---




